---
title: "Survival analysis project"
author: "DC, RD, RL, RR"
date: "2023-07-07"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1/ENVIRONMENT PREPARATION
First, let's install the libraries that will be required in our analysis

```{r}
library(tidyverse)
library(dplyr)
library(survival)
library(lubridate)
```

# 2/DATA PREPARATION
First, we need to specify the path where the dataset is located. You need to amend it with your own path
```{r}
setwd('C:/Users/romai/Documents/DSTI/21-Survival Analysis/UTMB')
data_utmb17 <- read_csv("utmb_2017.csv", col_names = TRUE)
head(data_utmb17)
```


Let's check if we get some problems during the data import
```{r}
problems(data_utmb17)
```


Let's have a quick look on the dataset.
What are the columns?

```{r}
colnames(data_utmb17)
```

Let's get a bit more details on columns (type, etc)
```{r}
str(data_utmb17)
```

First column seems useless (it looks like a row numbering)
```{r}
data_utmb17 <- data_utmb17[,-1] 
```

We can see that column (category) contains 2 interesting information: age category and gender. Therefore, we can create 2 new columns for gender & age 
In addition, we add a column "status" (1 = finisher; 0 = DNF / did not finish) based on the presence or not of a time in the column "Arrivée"
```{r}

data_utmb17 <-data_utmb17 |>
  mutate(gender = case_when(
    endsWith(category, " H") ~ "Male",
    endsWith(category, " F") ~ "Female"),
    age = substring(data_utmb17$category, first=1, last=2),
    status = case_when(time != 'NA' ~ 1, TRUE ~ 0),
    .after ="category")
```

We can observe that there is no column capturing the latest/highest time for all individuals. Column "Arrivée" (Arrival  <=> finish line) capture only finisher (status =1). Non-finisher individuals (status = 0) have only the last time corresponding to the time where they stop the race.
Therefore, we create a new column a new column "HighestTime" to capture the information about the time-to-event regardless the status.
```{r}
data_utmb17$highesttime <- apply(data_utmb17[11:35], 1, function(x) max(x, na.rm = TRUE))
data_utmb17<-data_utmb17|>
  mutate(highesttime = replace_na(highesttime, '00:00:00'))
```

Format of the newly-created column "highesttime" is character preventing to apply survival analysis. 
```{r}
str(data_utmb17$highesttime)
```

Therefore, we convert it in time format (expressed in seconds) creating a the final time column "timetoevent'
```{r}
data_utmb17$timetoevent<- lubridate::hms(data_utmb17$highesttime)
data_utmb17$timetoevent<- period_to_seconds(data_utmb17$timetoevent)
```

Then, we remove all intermediate checkpoints time that are not useful anymore for our analysis
```{r}
data_utmb17<- data_utmb17[,-11:-34]
colnames(data_utmb17)
```

We keep removing others useless columns
  * name
  * team : only few individuals show that the information
  * category : we split it in 2 new columns (gender and age)
  * nationality: removed because we don't have the information for all censored individuals
  * Arrivée (arrival): we capture it in the timetoevent column
  * highestime: not the appropriate format -> convert in time format (seconds) above

We keep only useful columns: bib (or ID), gender, age, status and timetoevent

```{r}
data_utmb17<-  data_utmb17[,-c(2,3,4,8,9,10)]
```


Then, we convert the age category (SE, V1, V2, V3, V4) in age range (in years) using the international age ranking for running trail
```{r}
colnames(data_utmb17)

age_range <- tibble('age' =
                      c("V1", "V2","V3", "V4", "SE"),
                    'age_range' = c("40-49", "50-59", "60-69", "70+", "23-39")
                    )
data_utmb17 <- data_utmb17 |> inner_join(age_range,by = "age")

#Move column "age_range" just after column "age"
data_utmb17 <- data_utmb17  %>% relocate("age_range", .after = "age")
```

```{r}
table(data_utmb17$age_range)
```
The 3 oldest categories contains few individuals compared to the 2 others. We could merge the 3 oldest range together.

```{r}
data_utmb17 ["age_range"] [data_utmb17 ["age_range"] == "60-69"]<- "50+"
data_utmb17 ["age_range"] [data_utmb17 ["age_range"] == "70+"]<- "50+"
data_utmb17 ["age_range"] [data_utmb17 ["age_range"] == "50-59"] <- "50+"
```

```{r}
table(data_utmb17$age_range)
```
```{r}
table(data_utmb17$age_range, data_utmb17$gender)
```

# 3/SURVIVAL ANALYSIS
## a/Global analysis
### Kaplan-Meier
```{r}

fit.KM <- survfit(Surv(timetoevent, status) ~ 1, data = data_utmb17)
fit.KM


plot(fit.KM, mark.time = TRUE,
     main = "Kaplan-Meier estimator",
     ylab = "Survival probability",
     xlab = "time (seconds)")

```

## b/ Group by AGE
### Kaplan-Meier
```{r}

fit.KMage <- survfit(Surv(timetoevent, status) ~ age_range, data = data_utmb17)
fit.KMage

plot(fit.KMage, col = 13:16)
legend("bottomleft", lty = 1, col = 13:16, legend = names(fit.KMage$strata))
```

### Log rank test
The logrank test is the most widely used method of comparing two or more survival curves

```{r}
diff.KMage <- survdiff(Surv(timetoevent, status) ~ age_range, data = data_utmb17)
diff.KMage

```
p-value = 7e-12  (<<0.05), we reject H0 => there exists at least a significant difference between 2 age range reinforcing the visual impression of a trend towards better survival (chance to finish the race) when the age is less advanced.


### Semi-parametric Cox regression
```{r}
cox.age<- coxph(Surv(timetoevent, status) ~ age_range, data = data_utmb17)
summary(cox.age)
```
The reference group is the youngest group (23-39). The Cox regression shows that the 2 other age groups are statistically significant compared to the reference (p<<0.05).
The impact of the age decrease the risk h of finishing the race by 0.8 and 0.6 (respectively for 40-49 and 50+) meaning that the youngest group has, respectively, 1.25 times and 1.66 times more chance to finish the race.


## c/ Group by GENDER
### Kaplan-Meier
```{r}
fit.KMgender <- survfit(Surv(timetoevent, status) ~ gender, data = data_utmb17)
fit.KMgender

plot(fit.KMgender, col = c("#FF3399", "#0066FF"),pch = 19)
legend("bottomleft", lty = 1, col = c("#FF3399", "#0066FF"), cex= 0.75, legend = names(fit.KMgender$strata))


```
### Log rank test by gender
```{r}
diff.KMgender <- survdiff(Surv(timetoevent, status) ~ gender, data = data_utmb17)
diff.KMgender

```
The p-value is large (p=0.1): the difference *is not* statistically significant.

As we can see on the KM curve, both curves are crossing twice. We can suspect an influence of the age. Let's now stratify on the age to see of we can observe a difference between gender

```{r}
diff.KMgender2 <- survdiff(Surv(timetoevent, status) ~ gender + strata(age_range), data = data_utmb17)
diff.KMgender2
```
The p-value decrease a bit (p=0.07 vs 0.1) that let us think about age influence but the difference is still not statistically significant between male and female.
We can note that the number of female is lower than male. We could increase the size of female sample to have a more balanced dataset to improve the analysis.

### Semi-parametric Cox regression
```{r}
cox.gender<- coxph(Surv(timetoevent, status) ~ gender + strata(age_range), data = data_utmb17)
summary(cox.gender)
```
The reference group is the female group. The Cox regression shows that the male group is not statistically significant compared to the reference confirming the log rank test analysis.
The impact of the gender increase the "risk" of finishing the race by 1.153 meaning that the female have 0.867 times more chance to finish the race than male.

## d/ Group by Age AND Gender
### Kaplan-Meier
```{r}
fit.KMage_gender <- survfit(Surv(timetoevent, status) ~ age_range + gender, data = data_utmb17)
fit.KMage_gender

plot(fit.KMage_gender, col = 1:9)
legend("bottomleft",lty = 1, col = 1:9, legend = names(fit.KMage_gender$strata), cex= 0.6, box.lty=1)
```

### Log rank test
```{r}
diff.KMage_gender1 <- survdiff(Surv(timetoevent, status) ~ gender + age_range , data = data_utmb17)
diff.KMage_gender1
```

### Semi-parametric Cox regression
without interaction btw age and sex 
```{r}
cox.age_gender1<- coxph(Surv(timetoevent, status) ~ gender + age_range, data = data_utmb17)
summary(cox.age_gender1)
```

with interaction btw age and sex (age:gender)
```{r}

cox.age_gender2<- coxph(Surv(timetoevent, status) ~ gender + age_range + age_range:gender, data = data_utmb17)
summary(cox.age_gender2)
```
The interaction age:gender is not statistically significant meaning that we can remove it from the model.
In fact, we observe same results as before meaning that only the covariate age have a significant impact of the survival.


## e/ Comparison & validation of the Cox models
### AIC 
Let's compare the different Cox models and see which is the "best" one using AIC:cox.age, cox.gender (with stratification on age), cox.age_gender1 (no interaction) and cox.age_gender2 (with interaction)

```{r}
fits<- list(MA = cox.age, MB = cox.gender, MC1 = cox.age_gender1, MC2=cox.age_gender2)
sapply(fits, AIC)
```
We can see that the best model (with lowest AIC) is the model considering the covariate "gender"(note: we did not observe a significant of gender on survival time!)

### Proportionality of hazards : complementary log-log plot
Let's consider the model with the covariate "age_range" for which we observed a significant difference between the youngest group and the 2 others

```{r}
plot(survfit(Surv(timetoevent, status) ~ age_range , data = data_utmb17),
     fun = "cloglog",
     col = 1:2)
```
In this figure we see that the different age_ranges are not parallel which indicates that the proportionality of hazards is not respected.
```{r}
plot(survfit(Surv(timetoevent, status) ~ gender , data = data_utmb17),
     fun = "cloglog",
     col = 1:2)
```
In this figure we see that the different gender are not parallel which indicates that the proportionality of hazards is not respected.

```{r}
plot(survfit(Surv(timetoevent, status) ~ gender + strata(age_range) , data = data_utmb17),
     fun = "cloglog",
     col = 1:2)
```
### Schoenfeld residuals plot
Let's consider the model with the covariate "age_range" for which we observed a significant difference between the youngest group and the 2 others.
```{r}
test.ph <- cox.zph(cox.age)
test.ph
plot(test.ph)
```
The output of the test is p < 0.05 which is statistically significant. Therefore, the proportional hazards assumption is not respected and is in accordance with the complementary log-log plot
```{r}
test.ph <- cox.zph(cox.gender)
test.ph
plot(test.ph)
```
The output of the test is p > 0.05 which is statistically non- significant. Therefore, the proportional hazards assumption is respected and is in accordance with the complementary log-log plot concerning the gender covariate
